---
title: "R Notebook 3: OLS Unbiasedness & Sampling Distribution ($E[\\hat{\\beta}] = \\beta$)"
author: "Nick Lim"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: readable
    highlight: tango
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8.5, fig.height = 5.2)
library(ggplot2)
library(dplyr)
library(purrr)
library(broom)

set.seed(42)
theme_set(theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    panel.border = element_rect(colour = "#cbd5e1", fill = NA, linewidth = 0.8)
  ))
```

# Introduction: What is Unbiasedness?

Under the Gauss-Markov assumptions, Ordinary Least Squares (OLS) estimators are **unbiased**, meaning their expected value across all possible random samples equals the true underlying population parameter:
$$E[\hat{\beta}_0] = \beta_0, \quad E[\hat{\beta}_1] = \beta_1$$

In this simulation notebook, we demonstrate what this theoretical expectation means practically:
1. We generate a true **Population of 1,000 points** from the linear process $Y = 1.0 + 2.0X + \mathcal{N}(0, 1)$ where $X \sim \text{Uniform}(0, 10)$.
2. We run **1,000 Monte Carlo loops**: in each loop, we draw a random subsample of $n = 30$ observations without replacement from the population, fit `lm(y ~ x)`, and record the estimated intercept ($\hat{\beta}_0$) and slope ($\hat{\beta}_1$).
3. We plot histograms of the 1,000 estimated coefficients to show that their average centers exactly on $\beta_0 = 1.0$ and $\beta_1 = 2.0$.

---

# 1. Generating the True Population ($N = 1,000$)

Let's simulate the underlying 1,000 population data points.

```{r pop-data}
N_pop <- 1000
pop_df <- tibble(
  id = 1:N_pop,
  x = runif(N_pop, min = 0, max = 10),
  y = 1.0 + 2.0 * x + rnorm(N_pop, mean = 0, sd = 1.0)
)

ggplot(pop_df, aes(x = x, y = y)) +
  geom_point(color = "#3b82f6", alpha = 0.5, size = 2) +
  geom_abline(intercept = 1.0, slope = 2.0, color = "#1e293b", linewidth = 1.5) +
  labs(
    title = "True Population Relationship (N = 1,000)",
    subtitle = "Solid black line indicates true population process: Y = 1 + 2X",
    x = "Predictor (X)",
    y = "Outcome (Y)"
  )
```

---

# 2. Monte Carlo Simulation: 1,000 Samples of $n = 30$

We now perform the sampling experiment using `purrr::map_dfr()` to collect intercept and slope estimates across 1,000 independent samples of size 30.

```{r mc-sim}
n_sims <- 1000
sample_size <- 30

sim_results <- map_dfr(1:n_sims, function(i) {
  sample_df <- pop_df %>% slice_sample(n = sample_size)
  mod <- lm(y ~ x, data = sample_df)
  tidy(mod) %>%
    select(term, estimate) %>%
    tidyr::pivot_wider(names_from = term, values_from = estimate) %>%
    rename(beta_0 = `(Intercept)`, beta_1 = x)
})

cat("=== Sampling Distribution Summary Table ===\n")
cat("True Intercept (beta_0)             : 1.0000\n")
cat("Mean Sample Intercept (mean(hat_b0)): ", round(mean(sim_results$beta_0), 4), "\n")
cat("-> Intercept Bias                   : ", round(mean(sim_results$beta_0) - 1.0, 4), "\n\n")

cat("True Slope (beta_1)                 : 2.0000\n")
cat("Mean Sample Slope (mean(hat_b1))    : ", round(mean(sim_results$beta_1), 4), "\n")
cat("-> Slope Bias                       : ", round(mean(sim_results$beta_1) - 2.0, 4), "\n")
```

---

# 3. Visualizing the Sampling Distributions

Let's plot side-by-side histograms of the 1,000 sample intercepts and slopes. We draw:
- **Dashed Black Line:** True population value ($\beta_0 = 1.0, \beta_1 = 2.0$)
- **Solid Red Line:** Average of our 1,000 estimated coefficients across all samples

```{r plot-histograms, fig.width=10, fig.height=5}
p1 <- ggplot(sim_results, aes(x = beta_0)) +
  geom_histogram(fill = "#f59e0b", color = "white", bins = 35, alpha = 0.75) +
  geom_vline(xintercept = 1.0, color = "#1e293b", linetype = "dashed", linewidth = 1.4, aes(color = "True Parameter (1.0)")) +
  geom_vline(xintercept = mean(sim_results$beta_0), color = "#dc2626", linewidth = 1.2, aes(color = "Sample Mean")) +
  labs(
    title = "Sampling Distribution of Intercept",
    subtitle = paste0("Mean estimate = ", round(mean(sim_results$beta_0), 3)),
    x = expression(Estimated ~ Intercept ~ (hat(beta)[0])),
    y = "Frequency"
  )

p2 <- ggplot(sim_results, aes(x = beta_1)) +
  geom_histogram(fill = "#16a34a", color = "white", bins = 35, alpha = 0.75) +
  geom_vline(xintercept = 2.0, color = "#1e293b", linetype = "dashed", linewidth = 1.4) +
  geom_vline(xintercept = mean(sim_results$beta_1), color = "#dc2626", linewidth = 1.2) +
  labs(
    title = "Sampling Distribution of Slope",
    subtitle = paste0("Mean estimate = ", round(mean(sim_results$beta_1), 3)),
    x = expression(Estimated ~ Slope ~ (hat(beta)[1])),
    y = "Frequency"
  )

library(patchwork)
if (requireNamespace("patchwork", quietly = TRUE)) {
  p1 + p2
} else {
  print(p1)
  print(p2)
}
```

---

# Key Takeaways

1. **Unbiasedness is an Aggregate Guarantee:** While any individual sample of $n=30$ might produce a slope like `1.82` or `2.18`, the average across all possible samples converges exactly to `2.0`.
2. **Gauss-Markov Efficiency:** Under constant error variance and zero conditional mean, OLS not only yields unbiased parameter estimates but also achieves the minimum variance among all linear unbiased estimators (**BLUE** property).
