---
title: "R Notebook 1: Cook's Distance & Leverage Diagnostics"
author: "Nick Lim"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: readable
    highlight: tango
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)
library(ggplot2)
library(dplyr)
library(broom)

set.seed(42)
theme_set(theme_minimal(base_size = 14))
```

# Introduction

In this notebook, we explore **Assumption 5 (Normality of Errors & Low Sample Sensitivity)** and **Outlier Influence**. While Ordinary Least Squares (OLS) does not strictly require normal errors to calculate unbiased slopes when sample sizes are large (thanks to the Central Limit Theorem), **small samples are acutely vulnerable to extreme outliers**. Because OLS minimizes the **squared** residuals, a single high-leverage observation can pull the entire regression line off course like a magnet.

Here, you will learn:
1. How to measure **Leverage** (`hatvalues`) and **Cook's Distance** ($D_i$).
2. How to visualize Cook's Distance against the critical threshold ($4/n$).
3. How removing or re-weighting influential points restores accurate parameter estimation.

---

# 1. Simulating Small Sample Data with an Outlier

Let's generate $n=14$ clean observations from $Y = 1.0 + 1.2X + \epsilon$, and then append a 15th observation that is an extreme leverage outlier at $X=12.0, Y=-3.0$.

```{r simulate-data}
n_clean <- 14
x_clean <- seq(1, 8, length.out = n_clean)
y_clean <- 1.0 + 1.2 * x_clean + rnorm(n_clean, mean = 0, sd = 0.5)

# Add extreme outlier
df <- tibble(
  id = 1:(n_clean + 1),
  x = c(x_clean, 12.0),
  y = c(y_clean, -3.0),
  type = c(rep("Standard Observation", n_clean), "Extreme Leverage Outlier")
)

# Fit OLS models
mod_clean <- lm(y ~ x, data = filter(df, type == "Standard Observation"))
mod_outlier <- lm(y ~ x, data = df)

cat("=== OLS Slopes Comparison ===\n")
cat("Clean Data Slope (True ~ 1.20):", round(coef(mod_clean)[2], 3), "\n")
cat("Outlier Contaminated Slope    :", round(coef(mod_outlier)[2], 3), "\n")
```

---

# 2. Visualizing the Magnetic Pull of Outliers

Let's plot both regression lines to observe how the single point drags down the slope.

```{r plot-regression}
x_grid <- seq(0, 13, length.out = 100)
df_lines <- bind_rows(
  tibble(x = x_grid, y = predict(mod_clean, newdata = data.frame(x = x_grid)), model = "Without Outlier (Clean)"),
  tibble(x = x_grid, y = predict(mod_outlier, newdata = data.frame(x = x_grid)), model = "With Outlier (Contaminated)")
)

ggplot(df, aes(x = x, y = y)) +
  geom_point(aes(color = type, size = type), alpha = 0.85) +
  geom_line(data = df_lines, aes(x = x, y = y, linetype = model, color = model), linewidth = 1.3) +
  scale_color_manual(values = c("Standard Observation" = "#2563eb", "Extreme Leverage Outlier" = "#dc2626",
                                "Without Outlier (Clean)" = "#16a34a", "With Outlier (Contaminated)" = "#dc2626")) +
  scale_size_manual(values = c("Standard Observation" = 3.5, "Extreme Leverage Outlier" = 6)) +
  labs(
    title = "The Magnetic Pull of Extreme Leverage Outliers",
    subtitle = "Notice how one observation tilts the slope from positive to negative/zero!",
    x = "Predictor (X)",
    y = "Outcome (Y)"
  ) +
  theme(legend.position = "bottom")
```

---

# 3. Cook's Distance Diagnostic ($D_i$)

How do we systematically flag points that exert undue influence? We use **Cook's Distance**, which measures how much all fitted values $\hat{Y}$ change when observation $i$ is removed from the dataset:

$$D_i = \frac{\sum_{j=1}^n (\hat{Y}_j - \hat{Y}_{j(i)})^2}{p \cdot s^2}$$

A common rule of thumb flags observations where $D_i > \frac{4}{n}$ as highly influential.

```{r cooks-distance}
df_aug <- augment(mod_outlier) %>%
  mutate(
    id = row_number(),
    threshold = 4 / n(),
    is_influential = .cooksd > threshold
  )

cat("Critical Cook's Distance Threshold (4/n):", round(4/nrow(df), 3), "\n\n")
print(select(df_aug, id, x, y, .hat, .cooksd, is_influential))
```

Let's plot Cook's Distance across all observation indices:

```{r plot-cooks}
ggplot(df_aug, aes(x = id, y = .cooksd)) +
  geom_segment(aes(xend = id, yend = 0, color = is_influential), linewidth = 1.2) +
  geom_point(aes(color = is_influential), size = 4) +
  geom_hline(aes(yintercept = threshold), linetype = "dashed", color = "#dc2626", linewidth = 1) +
  scale_color_manual(values = c("FALSE" = "#2563eb", "TRUE" = "#dc2626"), name = "Undue Leverage (D > 4/n)") +
  labs(
    title = "Cook's Distance Diagnostic Chart",
    subtitle = "Dashed red line indicates critical threshold (4/n)",
    x = "Observation Index (i)",
    y = "Cook's Distance (D_i)"
  ) +
  theme(legend.position = "top")
```

---

# Summary & Best Practices

1. **Check Cook's Distance (`.cooksd`) and Leverage (`.hat`)** whenever running OLS on small sample sizes ($n < 30$).
2. **Do not blindly delete outliers!** Investigate whether the outlier is a data entry/measurement error, or a genuine extreme phenomenon.
3. If genuine, consider using **Robust Regression** (e.g., `MASS::rlm()` which uses M-estimation) rather than standard OLS squared error minimization.
